Supplement: Potential and limits of site-specific substitution model inference.

### 2020-01-18 manuscript/supplement.tex
typos and streamlining. Move supp figures to supplement

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

@@

\begin{document}

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{xr}
\externaldocument{ms}
%\usepackage{cite}

@@

\begin{document}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}

@@

\clearpage

\section*{Supplementary figures}
\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_aa_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_aa_ratealpha1.5}.pdf}
	\caption{Accuracy of iterative estimation as a function of the total tree length, i.e., the expected number of state changes along the tree, for amino acid alphabets. (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the number of observable mutations.
	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000]$ sequences.
	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence.  }
	\label{fig:dressed_supp}
\end{figure*}

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/aa_length_n300}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/aa_depth_n300}.pdf}
	\caption{{\bf Skewed equilibrium concentration results in branch length under estimates -- amino acid alphabet}.
	Panel A shows the inferred average branch length as estimated by IQ-Tree and TreeTime as a function of the true average branch length. Panel B shows the results of the same optimization for the average root-to-tip distance, which is dominated by deep long branches which are more strongly affected by underestimation. Note that inferred site-specific models only partially ameliorate underestimation, see main text. }
	\label{fig:tree_length_supp}
\end{figure*}

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_gag_fitness_pc_0.010}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_nef_fitness_pc_0.010}.pdf}
	\caption{{\bf Intra-host vs cross-sectional mutation selection balance.}
	In analogy to Fig.~\ref{fig:HIV_fitness}, the panels A\&B show the ratio of in/out rates for consensus nucleotides acids along the \gene{gag} and \gene{nef} gene of HIV-1 subtype B vs of fitness costs of non-consensus states estimated from within-host mutation selection balance.
	The logarithm of the rate ratio is roughly linear in the logarithm of the fitness cost.}
	\label{fig:HIV_fitness_supp}
\end{figure*}

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/HIV_B_pol_alignment_div_pc_0.010_aa}.pdf}
	\caption{{\bf Diversity and GTR rate estimates.}
	Panel A\&B shows the ratio of in/out rates for consensus nucleotides/amino acids along the \gene{pol} of HIV-1 subtype B vs the diversity in columns of the corresponding alignment.}
	\label{fig:HIV_diversity_supp}
\end{figure*}
